---
title: Why is recursive sql so confusing?
author: Ted Farley
date: '2021-11-06'
slug: []
categories: []
tags: []
---
```{r setup, echo=F}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
# knitr::knit_engines$set(list(python = "/Users/edwardfarley/Library/r-miniconda/envs/r-reticulate/bin/python"))
knitr::knit_engines$set(py = function(options) {
  code <- paste(options$code, collapse = '\n')
  out  <- system2(
    'python', c('-c', shQuote(code)), stdout = TRUE
  )
  knitr::engine_output(options, code, out)
})
```

```{r, echo = F}
library(duckdb)
library("DBI")
con = dbConnect(duckdb::duckdb(), ":memory:")
```

Recursion is useful when trying to represent hierarchical relationships. The trouble with relational SQL is that is is difficult and confusing to represent hierarchies from rectangular relations-- hierarchies are best represented by trees.

## Distance from CEO problem

Below, I'll give an example of a very simple use-case for recursive sql: The depth from president problem[^1]. The problem is, given an unique `employee_id` and `manager_id` that relates to another employee can you tell how 'far removed' the given employee is from the CEO of the company. 

The example looks innocent enough until you consider that the solution should be able generalize to companies with arbitrary organizational structures. The generalized solution requires recursion. Below I'll give the solution to the problem in SQL and then in python. I'll explain why I think the SQL solution is more confusing to me than the python solution.





```{sql ddl, connection=con}
/*
Make the `employees` table.
I include titles here to make the setup more intuitive
*/

create table employees as 
  select 
    * 
  from (
  values 
  (1001,null,'CEO'),
  (2002,1001,'Director'),
  (3003,1001,'Office Manager'),
  (4004,2002,'Engineer'),
  (5005,2002,'Engineer'),
  (6006,2002,'Engineer')) employees(id, manager_id, title);
  
```

```{sql connection=con}

with recursive employees_cte as (

  select -- base-case select statement
    id,
    0 as depth
  from employees
  where title = 'CEO'

  union all 

  select 
  
    employees.id,
    employees_cte.depth + 1 -- only iterates when the next subordinate is found
  
  from employees
  inner join employees_cte on --implicit recursive call
        employees.manager_id = employees_cte.id -- defines the stopping condition
  
)

select 'Employee ' || id || ' depth is ' || depth as output from employees_cte;
```


### Lets compare this to python 

```{r retic, echo=F}
library(reticulate)
```




```{py}
ids=[1001, 2002, 3003, 4004, 5005, 6006]
mgr_ids=[None, 1001, 1001, 2002, 2002, 2002]


def getDistance(pos):
    
    # stopping condition
    if mgr_ids[pos] is None:
        return 0
    
    for i in range(0, len(ids)):
        if ids[i] == mgr_ids[pos]:
            
            # recursive step
            return 1 + getDistance(i)

for i in range(0,len(ids)):
    print(f'Employee {ids[i]}  depth is  {getDistance(i)}')
```




### The CTE calling itself

The first reason that I find SQL recursion confusing is because the recursive step happens in an inner join. When we *inner join* the `employees` table on `employees_cte` we are implicitly calling the CTE on itself.


```{sql connection = con, eval=F}
inner join employees_cte
```

In python this step is very explicit:

```{python, eval = F}
def getDistance(pos):
  ...
  ...
  return 1 + getDistance(i) # recursive step
```

### Stopping condition as a join

The second way that SQL obscures recursive operations is that the stopping condition is actually in the join predicate:


```{sql connection = con, eval=F}
employees.manager_id = counter.id
```


this is why you can only do inner join's with recursive CTE's. The recursive query needs the limitations of an *inner join* or else it will call itself infinitely. 

Compare this to the python code:

```{python, eval = F}
def getDistance(pos):
    
    # stopping condition
    if mgr_ids[pos] is None:
        return 0
    ...
    ...
```

it's much easier to spot the stopping condition in python because its very explicitly stated in an if-else statement.

## Conclusion

I've never used a recursive query to solve a problem at work but that's probably because I haven't known how to do them for very long! While I can't find any cases where I would need to use them, [here is at least one person](https://haughtcodeworks.com/blog/software-development/recursive-sql-queries-using-ctes/) that needed them! In this post, I have outlined how recursion can dramatically simplify a certain type of problem. I have also shown that some of the abstractions that make SQL so wonderful can actually make it's recursive functionality more difficult to understand. 






[^1]: I got it from Puzzle #2: Managers and Employees, [here](https://github.com/smpetersgithub/AdvancedSQLPuzzles/blob/main/Advanced%20SQL%20Puzzles/Advanced%20SQL%20Puzzles.pdf)








